---
title: "Climate Change in Europe"
author: "Alexa Fredston, PhD"
date: 'May 4, 2021'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE, 
                      warning = FALSE)
```

# Introduction

The goal of this tutorial is to demonstrate how R can be used to access, wrangle, and visualize data about climate change. 

We'll focus on the effects of XXXXX in Europe. 

# Load packages

This demonstration is intended to introduce you to climate data without delving into gridded datasets, spatial data analysis and visualization, etc. -- each a topic worthy of an entire tutorial! We're going to use some custom packages developed by [Mike Johnson](https://mikejohnson51.github.io/) to facilitate this. If you're interested in doing a more complete analysis of something climate-related, you'll need additional tools. 

If you haven't already, please install these packages:

```{r installation, eval=FALSE}
install.packages("tidyverse") # for data wrangling 
install.packages("lubridate") # for managing dates
install.packages("sf") # "simple features", a comprehensive spatial analysis package for R
install.packages("raster") # a set of GIS (geographic information systems) resources for handling raster (gridded) spatial data 
install.packages("rnaturalearth") # for map background
remotes::install_github("mikejohnson51/AOI")
remotes::install_github("mikejohnson51/climateR")
```

```{r packages}
library(tidyverse)
library(lubridate)
library(sf)
library(raster)
library(AOI) # "area of interest" 
library(climateR)
select <- dplyr::select
```

# Fetch data

Update the country of interest to anywhere you'd like to explore! I'm starting with the Netherlands.

```{r spatial data}
country_of_interest <- "Netherlands"

AOI <- aoi_get(country=country_of_interest)
```

This is a polygon object of class `sf`, meaning it was created with the `sf` package. The polygon looks like this: 

```{r AOIpolygon}
plot(AOI$geometry)
```

Confirm where the AOI is by mapping it with a custom function from this package.

```{r AOImap}
AOI %>% aoi_map(returnMap = T)
```

Now we need to download climate data from the same region. Below, we're fetching data on maximum temperatures from a dataset called TerraClim.

```{r climate data}
# param_meta$terraclim # run this to see all the variables in terraclim, and feel free to replace tmax with a different one! 
var_of_interest <- "tmax"
dat_raster <- getTerraClim(AOI, param = var_of_interest, startDate = "1980-01-01", endDate = "2015-12-31") # will take a few seconds to run 
# class(dat)
# class(dat$terraclim_tmax)
```

You can see that this returns a list object with only one element, which is a RasterStack. This represents a set of spatial objects of equal area with different values (here, temperature values over time) that are "stacked".

# Explore data

First, let's get the temperature data into a `sf` format, which allows us to both plot and analyze it as a spatial object, and access the values in a dataframe. 

```{r sf}
dat_sf <- dat_raster$terraclim_tmax %>% 
  as(., "SpatialPolygonsDataFrame") %>%  # necessary intermediate step for translating a raster into a polygon 
  st_as_sf() %>% # transform into sf object
  pivot_longer(-geometry, names_to="Date", values_to="Tmax") %>%  # tidy the dataframe part of it 
  mutate(Date = gsub("X", "", Date), # clean up formatting of date 
         Date = gsub("\\.", "-", Date), # need to "escape" the period
         Date = lubridate::parse_date_time(Date, orders = c("Y/m")), # convert to date format -- takes some extra work because date format is unusual
         coords = as.character(geometry)
  ) 

class(dat_sf)

class(dat_sf$Date)

head(dat_sf$Date)
```

How have temperatures changed over time?

```{r Tplot}
dat_sf %>% ggplot(aes(x=Date, y=Tmax)) +
  geom_line()
```

This isn't terribly informative, because we're mostly just looking at seasonal variation in temperature. Instead, let's plot the max temperature every year. 

```{r Tplot_year}
dat_sf %>% 
  mutate(year = lubridate::year(Date)) %>% 
  group_by(year) %>% 
  summarise(Tmax_yr = max(Tmax)) %>% 
  ggplot(aes(x=year, y=Tmax_yr)) +
  geom_line()
```

Of course, this isn't a very beautiful plot, and if we had more time we could work on making it a better visualization. But for now, let's just see if there appears to be an increase in maximum temperature over time. 

```{r Tplot_year_lm}
dat_sf %>% 
  mutate(year = lubridate::year(Date)) %>% 
  group_by(year) %>% 
  summarise(Tmax_yr = max(Tmax)) %>% 
  ggplot(aes(x=year, y=Tmax_yr)) +
  geom_line() +
  geom_smooth(method="lm")

```

This is the max temperature across the whole country; what if instead we looked at the max temperatures in every spatial point every year?

```{r Tplot_dots_lm}
dat_sf %>% 
  mutate(year = lubridate::year(Date)) %>% 
  group_by(year, coords) %>%  # now also grouping by coordinates
  summarise(Tmax_yr = max(Tmax)) %>% 
  ggplot(aes(x=year, y=Tmax_yr)) +
  geom_point() +
  geom_smooth(method="lm")

```

Just to confirm the increase is real, let's look at the results of a linear model fitted to this data.

```{r lm}
dat_sf %>% 
  mutate(year = lubridate::year(Date)) %>% 
  group_by(year, coords) %>% 
  summarise(Tmax_yr = max(Tmax)) %>% 
  lm(Tmax_yr ~ year, data=.) %>% 
  summary()
```

# Map data 

How has this played out over space? Let's map warming in this country for the most recent decade of data. 

```{r map}
europe <- rnaturalearth::ne_countries(continent = 'europe', returnclass = "sf") %>% 
  st_sf()

dat_sf %>%  
  mutate(year = lubridate::year(Date)) %>% 
  filter(year >= 2006) %>% 
  group_by(year, coords) %>% 
  mutate(Tmax_yr = max(Tmax)) %>% 
  select(year, Tmax_yr, geometry) %>% 
  distinct() %>% 
  ggplot() +
  geom_sf(data=europe, color="black") + 
  geom_sf(aes(geometry=geometry, fill=Tmax_yr, color=Tmax_yr)) +
  scale_color_gradient(low="yellow",high="red") + 
  scale_fill_gradient(low="yellow",high="red") + 
  scale_x_continuous(limits=c(1, 10)) +
  scale_y_continuous(limits=c(50, 55)) +
  facet_wrap(~year)
```

# Optional additional tasks 

Identify heatwave years -- the years with the greatest spikes in Tmax -- and map temperature in those years. 

Try changing countries and/or variables, and see if the visualizations need any adjustments. 

