---
title: "R for the Planet"
subtitle: "R-Ladies Amsterdam"
author: 
  - "Alexa Fredston, PhD"
date: 'May 4, 2021'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
seal: false
---
<style>
.title-slide {
vertical-align: bottom !important; 
text-align: left !important;
}
</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  fig.show = TRUE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(tweetrmd) 
style_duo_accent(
  primary_color = "#EBB59F",
  secondary_color = "#0B1B32",
  text_color = "#0B1B32",
  inverse_header_color = "#0B1B32",
  title_slide_text_color="#0B1B32",
  link_color = "#EBB59F",
  header_font_google = google_font("Cardo"),
  text_font_google = google_font("Work Sans"),
  code_font_google = google_font("Fira Mono")
)
```


# About our impact on the planet ... 

Yes, it's real.

Yes, it's bad.

Yes, it's us. 

No, it's not too late.

No, it isn't just climate change. 

--

### More on climate
[FAQs answered](https://www.nytimes.com/article/climate-change-global-warming-faq.html), a  [docu-series](https://www.amazon.com/Years-Living-Dangerously-Complete-Showtime/dp/B00NDOXDJU), a [YouTube channel](https://www.youtube.com/channel/UCi6RkdaEqgRVKi3AzidF4ow), a [bunch](https://www.criticalfrequency.org/drilled) [of](https://www.hottakepod.com/) [podcasts](https://www.degreespod.com/)

--

###Climate scientists to follow
[Jonathan Foley](https://twitter.com/GlobalEcoGuy), [Ed Hawkins](https://twitter.com/ed_hawkins), [Katharine Hayhoe](https://twitter.com/KHayhoe), [Leah Stokes](https://twitter.com/leahstokes)

---

# Why R?

Facilitates transparency and reproducibility

The R ecosystem (R + RStudio + GitHub + RMarkdown + ...) enables us to [do better science in less time](https://www.nature.com/articles/s41559-017-0160)

--

### Specific examples from my Twitter community 

```{r}
include_tweet("https://twitter.com/MikeTreglia/status/1382415509724004355?s=20")
```

---

# Why R?

Facilitates transparency and reproducibility

The R ecosystem (R + RStudio + GitHub + RMarkdown + ...) enables us to [do better science in less time](https://www.nature.com/articles/s41559-017-0160)

### Specific examples from my Twitter community 

```{r}
include_tweet("https://twitter.com/jebyrnes/status/1382377335387783174")
```

---

# Why R?

Facilitates transparency and reproducibility

The R ecosystem (R + RStudio + GitHub + RMarkdown + ...) enables us to [do better science in less time](https://www.nature.com/articles/s41559-017-0160)

### Specific examples from my Twitter community 

```{r}
include_tweet("https://twitter.com/jebyrnes/status/1382377335387783174")
```

---


# Why R?

Facilitates transparency and reproducibility

The R ecosystem (R + RStudio + GitHub + RMarkdown + ...) enables us to [do better science in less time](https://www.nature.com/articles/s41559-017-0160)

### Specific examples from my Twitter community 

```{r}
include_tweet("https://twitter.com/wormmaps/status/1382417628484997120?s=20")
```

--

```{r}
include_tweet("https://twitter.com/mdsumner/status/1382561385003974657?s=20")
```


---

# Why R?

Facilitates transparency and reproducibility

The R ecosystem (R + RStudio + GitHub + RMarkdown + ...) enables us to [do better science in less time](https://www.nature.com/articles/s41559-017-0160)

### Specific examples from my Twitter community 

```{r}
include_tweet("https://twitter.com/AdamClarkSmith/status/1382516148927918084")
```

---

# How environmental data scientists leverage R 

1. Mapping environmental health risks: redlining in California [code](https://github.com/daltare/Redline-Mapping), [app](https://cawaterdatadive.shinyapps.io/Redline-Mapping/)
1. Assessing the effects of marine protected areas: Ovando *et al.* 2021 [preprint](https://figshare.com/articles/preprint/Assessing_the_Population-level_Conservation_Effects_of_Marine_Protected_Areas/12715409), [code](https://github.com/DanOvando/population-effects-of-mpas)
1. Facilitating methodological transparency: The Ocean Health Index [Northeast Dashboard](https://ohi-northeast.shinyapps.io/ne-dashboard/)
1. Tracking marine species responses to climate change: [OceanAdapt](https://github.com/pinskylab/OceanAdapt)

---

# R, the planet, and you

Data about issues that matter to you is [at your fingertips](https://ropensci.org/packages/data-access/)...

--

... **But use it with caution!**

Domain knowledge matters

We're friendly. Just ask!

---

# Demo 


---
class: center, middle

# Thanks!

Slides created via the R packages:

[**xaringan**](https://github.com/yihui/xaringan)<br>
[gadenbuie/xaringanthemer](https://github.com/gadenbuie/xaringanthemer)

## Contact Information

[www.alexafredston.com](https://www.alexafredston.com/)<br>
[fredston@rutgers.edu](mailto:fredston@rutgers.edu)<br>
[@AFredston](https://twitter.com/AFredston)
